The tracing of trade

### 2021-02-10 Manuscript/Manuscript.tex
revision

\institute{Mar\'ia Coto-Sarmiento \at Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, Universitat de Barcelona, Barcelona, Spain
\email{mcotsar@gmail.com}            
\emph{mcotsar@gmail.com} 
%of F. Author  %  if needed
           \and
\\
Xavier Rubio-Campillo \at
School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK \at
DIDPATRI, Universitat de Barcelona, Passeig de la Vall d'Hebr\'on, 171, Barcelona, Spain             
}

@@

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade in \textit{Baetica} province (currently Andalusia). \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market.

Our case of study has been focused on amphoric production from 1st to 3rd centuries AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Amphoric stamps are used to identify the presence of different groups that might share similar stamps.

To achieve this goal, we analyse a set of stamps from two types of centres (production and consumption): 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} (Inferior and Superior) and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance. The analysis explores how the quantitative approach provides a useful tool for the interpretation of the economic processes. Finally, results pretend to highlight the organisation of Baetican olive oil production in the Roman Empire linked to the differences observed in the archaeological evidence.

%Insert your abstract here. Include keywords, PACS and mathematical subject classification numbers as needed.
\keywords{Roman Empire \and amphora production \and Dressel 20 \and dissimilarity index \and Roman provinces}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

The intensification of long-range trade was one of the most important traits of the economy developed during the Roman Empire. The development of an extensive road network increased the connectivity between inland communities, but most shipping continued being based on maritime routes particularly in the Mediterranean basin~\citep{temin_market_2001,bevan_mediterranean_2014}. The intensity of this trade between distant regions can be observed both in archaeological and written evidence~\citep{rodriguez_baetican_1998}.

The Empire developed a series of structures to support and organise this long-range trade and specialised entire regions to massively produce specific goods. An example of this process was the province of \textit{Baetica} (currently Andalusia, southern Spain) which became an important olive oil production centre during the Roman Empire. Olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which was distributed through amphorae shipped via maritime routes to all the provinces and particularly to Italy and the hundreds of garrisons that the Roman army deployed along its borders~\citep{blazquez_exportacion_1980}.

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. 
However, the lack of written records has not made possible to detect any indication of patterns in the olive oil market.

Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was amphora production organised during the Roman Empire? Can we detect economic dynamics between provinces by analysing the distribution of amphorae?

Advances in Roman studies have currently led to an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.

The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of Roman production
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

This work focuses on the debate about the dynamics of the distribution of olive oil market in Roman provinces. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in the provinces.

The finding of amphoric stamps in different provinces has allowed us to know the importance of the distribution of olive oil within the Empire. Despite having archaeological evidence, no consensus has been reached: How was the economic relation between production and consumption centres? Did they follow an economic pattern to distribute olive oil in those provinces?

This paper aims to study the olive oil market connection between provinces by calculating the similarity of amphoric stamps. Specifically, our work aims to detect regional processes focused on a commercial product (i.e. olive oil) from a specific province \citep{isaksen_network_2006}.

%We want to understand the pattern of olive oil production linked to amphora workshops and stamps used to mark them. 
%We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

Two case studies have been studied in order to analyse the relation between production centres (\textit{Baetica}) and consumption centres (\textit{Britannia} and \textit{Germania}) (see Figure~\ref{general}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{general_map}
\caption{Overview of the sites analysed in this paper. The colour defines the three different regions under study (\textit{Baetica} as producer; \textit{Germania} and \textit{Britannia} as consumers) while the size of each dot highlights the number of Dressel 20 amphoric stamps found on each of these sites}

@@

In the case of \textit{Baetica} province, we want to identify the role of the stamps in the organisation of the workshop; in Roman provinces, our aim is detecting groups of stamps concentrated in an area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could be identified by different aspects: a) correlation between spatial distance and the sites based on the idea that closer workshops concentrate similar amphoric stamps in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific province.

@@

Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of amphoric stamps, b) stamps located in close workshops share similar traits and c) low mobility of amphoric stamps to other regions: stamps always stay in the same region.

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. Quantitative approach provides one way to measure the similarity between amphoric stamps. Here, we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance.

@@

For example, if workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

The paper addresses these questions as follows: the next section introduces the historical context. Section two displays the dataset and the methods used for the analysis. Section three presents the results and the last section shows the discussion and the main conclusions of this work.

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increased demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third century AD~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were built near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).The proximity distance between workshops was relatively close which allowed for more efficient distribution.

This strategic location allowed the transport of olive oil through riverine shipping towards the maritime routes that connected \textit{Baetica} with Mediterranean and Atlantic trade routes towards the rest of the Empire~\citep{garcia_vargas_enrique_formal_2010}.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity \citep{remesal_anforas_2004}. Thus, the specialisation of this production made amphorae with apparently few changes in their forms, reaching a high degree of standardisation.

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 type amphorae found across the Roman provinces \citep{dressel_ricerche_1878,
millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.

They have a recognisable globular form with a short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

Dressel 20 amphorae display a lot of evidence, but some of it is still hard to interpret. A large percentage of Dressel 20 were marked with stamps, while they could also be inked with \textit{tituli picti} or incised with \textit{graffiti}. The interpretation for most of these inscriptions is still open due to the fragmentation of the material and the small sample size of well-preserved elements~\citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

\section{A potential indicator of Roman economy: Dressel 20 stamps}
\label{sec:2}

Dressel 20 was the amphora most stamped during the Roman Empire ~\citep[18]{millet_anforas_1998}. The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason, there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{dressel_ricerche_1878,
rodriguez_economioleicola_1977,
chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were mostly marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure}

Most stamps display a large activity of production that can be difficult to date using an accurate chronology. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
Despite \textit{tituli picti} contain important information about chronology, they have been still found in few quantities mostly concentrated in an only settlement such as \textit{Monte Testaccio}. In addition, the chronology of stamps could be also biased taking into account that amphorae were deposited after being produced and consumed.

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps had not a unique criterion. They showed a code with different letters and forms. Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to initial forms of a name but they can also appear complete.

Scholars agree that stamps are some type of identification mark, but there is no consensus on the meaning of these marks~\citep{rodriguez_baetican_1998}.

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}.

Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop?

@@

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes. This effectively led to a gradual change in the economic and social structures when the trade networks that supplied the Roman army became more complex.

Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}

However, some hypotheses argue elsewhere that this same \textit{praefectura annonae} could also have organised the supply of additional goods such as olive oil to the Roman legions  \citep{remesal_annona_1986,
remesal_annona_1990}

Olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in sites nearby military garrisons suggests the existence of diverse supply dynamics as there were trade processes flowing from military sites to civil settlements~\citep{remesal_annona_1986, carreras_britannia_1998}.

An explanation for the interest to export olive oil for the military garrison in the Roman provinces is still unclear. The origin of this exportation may be related to two aspects: a) consumption based on cultural reasons (i.e. social identification or habit  by a group of consumers) and b) economical and political reasons to create a redistributive system mechanism to supply military camps \citep[69-70]{carreras_britannia_1998}.

The Roman conquest in \textit{Britannia} and \textit{Germania} was a new opening to foreign trade in order to supply the military campaign. High demand for olive oil in both provinces stimulated an important trade network involving different agents and the Empire itself. \textit{Baetica} became a strategic province for the production and distribution in both provinces since it can be detected by the high presence of amphoric stamps found in civil and military areas during this period.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced on the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}.

The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. The demand for olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply.

This fact will have a particular intensity within sites close to the Hadrian Wall's garrisons. Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}.

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,
funari_corpus_1996,
carreras_britannia_1998,carreras_abastecimiento_2003}.

This intense consumption of olive oil also indicates a possible provincial structure designed to organise the supply of olive oil to military camps.

There are no written records explaining how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep{funari_economic_2005,orengo_seeds_2016}.

The consumption of olive oil would experience a progressive slowdown from the third century A.D. onwards. This date matches a change in market strategy in the Empire. This gradual decrease on \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by other types of amphorae from other places \citep{rodriguez1991aceite,millet_anforas_1998}.

@@

Roman studies exploring the presence of Dressel 20 amphorae in the provinces Inferior and Superior during Roman Empire have not had the same impact as the rest of the provinces \citep[293]{remesal_baetica_2002}.

The massive importation of Dressel 20 in both \textit{Germania} provinces could be explained due to the archaeological sources found mostly in German \textit{limes}. 
By contrast, it is still unknown if trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}.

\textit{Germania} presents a similar introduction pattern of olive oil than the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region while a majority of recovered Dressel 20 amphorae are located in the military sites that formed the German \textit{limes} \citep{remesal_germaniaengl_2002}.

The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a major concentration at the German \textit{limes}.

It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons \citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence related to the meaning of the amphoric stamps.

Previous studies suggested that the supply to the German \textit{limes} was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia})~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

@@

The goal of this study is exploring the effect of the production patterns between different centres. We are especially interested in identifying links between production and consumption centres by using amphoric stamps.

To study the amphoric stamps, we used the data collected in the CEIPAC database of amphoric epigraphy from different places~\citep{remesal_centro_2015} (see the database here \url{http://romanopendata.eu}). The CEIPAC dataset contains over 50.000 epigraphy records found on different types of amphorae, mostly from \textit{Monte Testaccio}. Most stamps used in the analysis did not have a specific chronology or were not available in the dataset. This fact did that we had to discard the possibility to analyse a distribution of stamps per chronologies.

In order to enable reproducibility of the results, the selected datasets and source codes are available on the following GitHub repository: \url{https://github.com/Mcotsar/Baetica\_stamps}.

@@

\subsection{Measuring the dissimilarity}
\label{sec:5}

>>>>

\institute{Mar\'ia Coto-Sarmiento \at Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, Universitat de Barcelona, Barcelona, Spain
\email{mcotsar@gmail.com}            
\emph{mcotsar@gmail.com} 
%of F. Author  %  if needed
           \and
\\
Xavier Rubio-Campillo \at
DIDPATRI, Universitat de Barcelona, Passeig de la Vall d'Hebr\'on, 171, Barcelona, Spain
School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK \at
}

@@

The aim of this study is to explore the trade dynamics involved on the large-scale distribution of olive oil produced in \textit{Baetica} province (currently Andalusia). \textit{Baetica} became an important production and distribution centre during the Roman Empire and its goods have been found in thousands of sites, but it is still unclear how exactly this large scale production was organised.

We explore here connections between production and consumption sites during the period of most intense activity (1st to 3rd centuries AD). Specifically, we use amphoric stamps as proxies to identify the workshops where amphorae found on consumer sites were produced.

To achieve this goal, we analyse a dataset of stamps found in two types of sites: 1) workshops from \textit{Baetica} province and 2) consumption centres located in two provinces \textit{Germania} (Inferior and Superior) and \textit{Britannia}. This dataset is used to detect links between specific sites that may share a large percentage of common stamps. The analysis is performed using quantitative methods borrowed from Ecology that allows us to group sites based on the amount of stamps shared between them. Results highlight the structure of economic processes of the Roman Empire using the archaeological evidence of olive oil large-scale distribution.

%Insert your abstract here. Include keywords, PACS and mathematical subject classification numbers as needed.
\keywords{Roman Empire \and amphora production \and Dressel 20 \and similarity index \and Roman provinces}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

The intensification of long-range trade was one of the most important traits of the economy developed during the Roman Empire. The development of an extensive road network increased the connectivity between inland communities while maritime and riverine shipping continued to be the most common trade methods, particularly in the Mediterranean basin~\citep{temin_market_2001,bevan_mediterranean_2014}. The intensity of this long-range trade between distant regions is observed both in archaeological and textual evidence~\citep{rodriguez_baetican_1998}.

The Empire developed a series of structures to support and organise this long-range trade and specialised entire regions to produce specific goods at a massive scale. The province of \textit{Baetica} (currently Andalusia, southern Spain) is an example of this specialization process. \textit{Baetica} became one of the most important olive oil production centres during the Roman Empire; olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which needed to be distributed to cities across thousands of kilometers. This large-scale trade was organised via the shipment of millions of amphorae using maritime and riverine routes to all the provinces and particularly to Italy and the military garrisons guarding the borders of the Empire~\citep{blazquez_exportacion_1980}.

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, duncan1982economy, Chic_hispania_1997,millet_anforas_1998}. However, the complexity and sheer scale of the process makes its understanding challenging despite several archaeologicale xcavations and the existence of some written records discussing its structure. Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was this massive amphora production organised? Can we detect the economic ties between provinces? Was this distribution centralized or each region organised it differently?

Recent advances in Roman studies have currently led to an environment with various interpretations on these commercial dynamics~\citep{temin_economy_2006,quantifyingwilson2009}. The application of diverse novel quantitative approaches to new accessible datasets has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the dynamics of Roman production\citep{brughmans_roman_2016,orengo_seeds_2016,bayesian_2018,coto-sarmiento_identifying_2018,rubio-campillo_ecology_2018}. This work aligns with these efforts and focuses on the distribution of olive oil towards Roman provinces in the borders. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in these provinces.

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence there is still no consensus on the structure that generated this pattern; how was the economic relation between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?

This paper aims to explore these questions regarding the olive oil market by identifying links between productionn centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

%We want to understand the pattern of olive oil production linked to amphora workshops and stamps used to mark them. 
%We focus here on exploring the economic relation between stamps and amphora production and distribution centres. 
This relation between the production centre (i.e. \textit{Baetica}) and consumption sites have been studied here for two case studies: \textit{Britannia} and \textit{Germania} (see Figure~\ref{general}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{general_map}
\caption{Sites analysed in this paper. The colour defines the three different regions under study (\textit{Baetica} (yellow) as producer; \textit{Germania} (blue) and \textit{Britannia} (green) as consumers) while the size of each dot is proportional to the number of Dressel 20 amphoric stamps found on each site}

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas : a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be found on specific provinces because they were shipped together.

@@

We proposed three working hypotheses based on these ideas: a) we can identify a correlation between spatial distance and the distribution of amphoric stamps, b) stamp sets found in nearby workshops should be more similar and c) there is a low mobility of amphoric stamps from consumption centres to other regions (i.e. stamps always stay in the same region).

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centre based on the amphoric stamps found on each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to stablish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

@@

The paper discusses these questions, the analysis and results as follows: next section introduces the historical context. Section two defines the dataset and the methods used for the analysis. Section three presents the results and section 4 discusses them; the paper finishes with some concluding remarks on the research performed.

@@

The \textit{Baetican} province was one of the most suitable regions to tackle the challenge of the huge demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive specialized infrastructure of olive oil production was gradually deployed. This production of olive oil had the largest peark from the first to the third century AD~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were built near a landscape covered by olive trees. These workshops made the amphorae which were filled with olive oil from the presses; they were built at the banks of the rivers Guadalquivir and Genil (see Fig.\ref{workshop}).

This proximity between production area, olive processing infrastructure, amphorae production and navigable rivers made the entire process extremely efficient. The strategic location allowed the transport of olive oil through riverine shipping towards the maritime routes that connected \textit{Baetica} with Mediterranean and Atlantic trade routes towards the rest of the Empire~\citep{garcia_vargas_enrique_formal_2010}.

@@

The chronology of the Baetican workshops is typically defined between the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Archaeological evidence suggests that this highly specialised production was developed across a long activity period\citep{remesal_anforas_2004}, but at the same time the process seemed to have been planned at some level as is observed by the relatively minor changes in amphoric typologies and the high degree of standardisation on the amphorae found across at least 50 workshops.

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 type amphorae found in all the Empire and particularly in the Western provinces~\citep{dressel_ricerche_1878,millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements~\citep{berni_millet_epigrafianforica_2008}. They have a recognisable globular form with a short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

Dressel 20 amphorae display a lot of interesting information beyond their shape, even though some of it is hard to interpret. A large percentage of Dressel 20 were marked with stamps, while some of them were also inked with \textit{tituli picti} or incised with \textit{graffiti}. The interpretation for most of these inscriptions is still open due to the fragmentation of the material, the lack of written records clarifying its use and for \textit{tituli} and \textit{graffiti} a small sample size of well-preserved elements~\citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

\section{Amphoric stamps as a proxy for economic structures}
\label{sec:2}

Dressel 20 was the amphora most stamped during the Roman Empire ~\citep[18]{millet_anforas_1998}. The large sample size of recovered Dressel 20 stamps is distributed across all Western provinces of the Roman Empire. This wide distribution is observed in the large volume of publications discussing the origin, use, chronology and spatial distribution of the stamped codes~\citep{dressel_ricerche_1878,rodriguez_economioleicola_1977,chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure}

Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found on nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the diposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they could also be imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps was not defined by a single criteria, but they often displayed a code composed by letters and additional symbols. One of the most common forms is a code of three letters showing a \textit{Tria Nomina}~\citep{berni_millet_amphora_1996}; in this case the three letters are the initials of a Roman name. In some cases the full name may appear, while the meaning of other stamps is difficult to interpret~\citep{rodriguez_baetican_1998}.

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. Despite the unsolved question of ownership some interesting ideas can be explored based on the finding of similar stamps across the Empire. Observable patterns can be used to explore how this production was organised and whether it is possible to identify specific production processes fdor olive oil trade. Our questions here will be focused on the distribution of amphoric stamps; were different stamps shipped through different distribution routes? Were specific stamp codes used in different workshops?

@@

The expansion of the Roman Empire to increasingly distant regions required an effort to intensify trade activity through Mediterranean and Atlantic routes. These needs effectively led to a gradual change in the economic and social structures as the trade networks that supplied the provinces and particularly the Roman army became more complex. An interesting example of this system is the figure of the \textit{praefectura annonae} which was created during Augustus's administration to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}.

Some authors argue elsewhere that this same \textit{praefectura annonae} could also have organised the supply of additional goods such as olive oil to the Roman legions~\citep{remesal_annona_1986,remesal_annona_1990}. From this perspective, olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in civilian sites close to the military garrisons also suggests the existence of diverse supply dynamics involving systematic trade between military sites to civil settlements~\citep{remesal_annona_1986, carreras_britannia_1998}. An explanation for the interest for exporting olive oil to the military garrison in the Roman provinces is still unclear. The reasons behind this effort to supply olive oil to all soldiers may be related to two aspects: a) consumption based on cultural reasons (i.e. social identification or habit  by a group of consumers) and b) economical and political reasons to create a redistributive system mechanism to supply military camps~\citep[69-70]{carreras_britannia_1998}.

The Roman conquest of \textit{Britannia} and \textit{Germania} was a new opportunity to expand economical activity to support military efforts. High demand for olive oil in both provinces stimulated an important trade network involving different agents and the Empire itself. \textit{Baetica} became a strategic province for the production and distribution in both provinces as can be observed by the large volume of Dressel 20 amphora vessels found both in civil and military sites of these areas during and after the conquest.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced in the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Archaeological records suggest that this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D. onwards \citep{peacock_amphorae_1991,carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exports linked to the arrival of legions caused by different military campaigns~\citep[161]{monfort_britanniaen_1998}.

The increase of Dressel 20 exports created an important commercial network for exchanges. The demand for olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply. This economic relation will be particularly intense for sites close to Hadrian Wall's garrisons. Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}.  This presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,funari_corpus_1996,carreras_britannia_1998,carreras_abastecimiento_2003}. This intense consumption of olive oil also suggests the existance of provincial structures created to organise the supply of olive oil across military camps. There are no written records detailing how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep{funari_economic_2005,orengo_seeds_2016}.

The consumption of Baetican olive oil would experience a progressive slowdown from the third century A.D. onwards. This date matches a change in market strategy in the Empire. This gradual decrease on \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by other types of amphorae from other places and particularly the African provinces~\citep{rodriguez1991aceite,millet_anforas_1998}.

@@

The distribution of Dressel 20 amphorae in the provinces of Germania Inferior and Germania Superior is not as well interpreted as the rest of provinces discussed in this work~\citep[293]{remesal_baetica_2002}. It is clear that there was a massive import of Dressel 20 in both provinces as is observed in archaeological excavations across the German \textit{limes}. However, it is still unknown to what extent trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}.

In any case, the introduction of olive oil in \textit{Germania} shows similar patterns than the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region, while a majority of recovered Dressel 20 amphorae are located in the military sites built over the German \textit{limes} \citep{remesal_germaniaengl_2002}. The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a major concentration at the German \textit{limes}. It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons~\citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence clarifying the meaning of the amphoric stamps.

Some studies assumed that Roman naval technology was not suited for oceanic transport and for this reason defined riverine shipping across Rhône and Rhine rivers as the basic supply route for the German \textit{limes}. However, more recent large-scale computational analysis as well as new excavations on the Iberian Atlantic coast suggest that the Atlantic route, which was far more efficient, may have played a major role on the supply of \textit{Germania} and \textit{Britannia}~\citep{rubio-campillo_ecology_2018}.

@@

The goal of this study is to explore these economic and long-range trade dynamics using amphoric stamps as proxies for common supply networks. We are especially interested on identifying links between Baetican production centres and consumption centres in the different \textit{limes}.

For this analysis we used data collected in the CEIPAC database of amphoric epigraphy~\citep{remesal_centro_2015}. The CEIPAC dataset contains over 50.000 epigraphic records found on different types of amphorae (the database is openly accessible from \url{http://romanopendata.eu}). A large percentage of these records come from from \textit{Monte Testaccio}, but the rest is distributed mainly across the Western provinces of the Empire. It is worth mentioning that the database does not provide temporal information due to the reasons previously discussed, and for this reason this work can only analyze the spatial distributions within a coarse chronology of 2 centuries (3rd-1st AD).

In order to enable reproducibility of the results, this dataset as well as the source code are available on the following GitHub repository: \url{https://github.com/Mcotsar/Baetica\_stamps}.

@@

\subsection{Quantifying similarity}
\label{sec:5}
